Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method

### 2017-05-25 HinesHetlandGPR.tex
added more

\documentclass[10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\title{Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method}
\author{Trever T. Hines and Eric A. Hetland}

% OUTLINE
% INTRODUCTION
% - Why we care about strain in the crust
%   - Knowing long term strain rates tells us where we can expect seismicity
%
% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processese
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
%
% APPLICATION TO CASCADIA SSE
% - SSE background
% - REML results
%   - noise 
%   - signal for IBM, SE, and WEN

@@

>>>>

\documentclass[10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{bm}
\title{Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method}
\author{Trever T. Hines and Eric A. Hetland}

% OUTLINE
% INTRODUCTION
% - Why we care about strain in the crust
%   - Knowing long term strain rates tells us where we can expect seismicity
%
% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processes. (include seasonal and spherical)
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
% - Numerical notes
%   - partition inverse and solve with CHOLMOD
%
% APPLICATION TO CASCADIA SSE
% - SSE background
% - REML results
%   - noise 
%   - signal for IBM, SE, and WEN
%   - Plot all covariances functions

@@

In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(x,t)$. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(x,t)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\mathbf{u}$ is a three-dimensional (two spatial dimensions and time) Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(x,t) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(x,t;x',t')$ is a covariance function indicating how we expect $u_i(x,t)$ to covary with $u_i(x',t')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}(x,t;x',t') = X(x,x')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

\subsection{Spatial and temporal covariance functions}
In this section, we list several covariance functions which are considered in this paper or could be of use in future studies. For see \citet{Rasmussen2006} for additional information.

The squared exponential covariance function,
\begin{equation}\label{eq:SE}
C(x,x') = \alpha^2 \exp\left(\frac{-||x - x'||_2^2}{2 \beta^2}\right),
\end{equation}
is commonly used in Kriging \citep[e.g,][]{Cressie1992} and Gaussian process regression \citep[e.g.,][]{Rasmussen2006}.  The squared exponential is a valid (i.e. positive definite) covariance function for any number of spatial dimensions, and it describes an isotropic Gaussian processes with realizations that are infinitely differentiable.  \citet{Kato1998} demonstrated that eq. (\ref{eq:SE}) is an appropriate covariance model for describing long-term tectonic strain rates.

The Wendland class of covariance functions \citep{Wendland2005} are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. Wendland covariance functions have compact support, which is a particularly useful property because we can exploit the sparsity of the the corresponding covariance matrices. In this study we use Wendland covariance functions to describe the temporal component of transient displacements. Therefore, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
C(x,x') = \alpha^2\left(1 - \frac{|x - x'|}{\beta}\right)^3_+ \left(\frac{3|x - x'|}{\beta} + 1\right), \ \ \ x \in \mathbb{R}^1,
\end{equation}
where
\begin{equation}
(x)_+ = 
\begin{cases}
x, \ \ \ x > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

They are positive definite for a finite dimensional space and are finitely differentiable.  If the length-scale is sufficiently small then we can take advantage of the sparsity of the covariance matrices in Section X. This turns out to significantly increase scale of the problems which can be addressed in this analysis.

FOGM solves the differential equation .... Has been used by Langbein.

Valid in ND

Brownian Motion is the solution to the differential equation .... Has been used by Langbein to describe noise
Valid in 1D

Integrated Brownian motion solves ... . It has been used by Segall to describe slip, Ohatani, Murray, 
Valid in 1D

We also want to give notable mention to the Periodic covariance function

\subsection{Constraining displacements and strain with GNSS data}
We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS data, $\bm{d}_*$, as the realization of a random vector $\bm{d} = [d_{ij}]^T$, where $d_{ij}$ describes observations at position $x_i$ and time $t_j$. We assume that the observations can be described as   
\begin{align}\label{eq:Data}
d_{ij} = &u(x_i,t_j) + \epsilon(x_i,t_j) + a_{i} + b_{i} t_j + \\
         &c_{1i}\sin(2 \pi t_j) + c_{2i}\cos(2 \pi t_j) + c_{3i}\sin(4 \pi t_j) + c_{4i}\cos(4 \pi t_j), 
\end{align}
where $a_i$ is an offset that is unique for each GNSS monument, $b_i$ is the steady rate of tectonic deformation at $x_i$, the sinusoids describe seasonal deformation (using units of years for $t$), and $\epsilon$ is a Gaussian process for noise that cannot be described parametrically. $\epsilon$ can consist of white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}.  For now, we describe $\epsilon$ generally as $\epsilon \sim \mathcal{N}(0,C_d)$. We consider the coefficients $a_{i}$, $b_{i}$, and $c_{kj}$, to be unknown and uncorrelated random variables. Of course, the tectonic deformation, $b_i$, is in reality spatially correlated and we could invoke tectonic model to form a prior on $b_i$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $b_i$ for each station and there is no need to incorporate a prior. Likewise, the seasonal coefficients may be spatially correlated as suggested by Langbein2008?. It may be worth exploring and exploiting such a correlation in a future study.

Optimal Hyperparameters

Conditioning the GP

Differentiation, scaling and addition to get strain
